Errata

### 2025-04-04 tex/Manuscript-1.tex
Refine methodology and results discussion sections for clarity and accuracy
- Updated the methodology section to enhance the description of microbial community data analysis techniques, ensuring consistency in terminology and clarity in the explanation of statistical methods.
- Revised the results discussion to improve the flow of information regarding biocrusts' impact on soil aggregate stability, emphasizing their role in arid ecosystems and correcting citation formatting.
- Adjusted references to ensure proper citation style and accuracy throughout the text.

    $$\Delta MWD = \frac{\sum_{i=1}^{n} X_i*W_i}{\sum_{i=1}^{n} W_i}$$
    where $W_i$ is the corrected mass proportion of stable aggregate fraction $i$ in the total 2–30 mm aggregate fractions, and $X_i$ is the mean diameter of stable aggregate fraction $i$. and $n$ is the number of particle fractions.

@@

The influence of the climatic gradient (study site) and biocrust presence on physicochemical soil parameters and aggregate stability in 40 plots (4 study sites, 2 biocrust treatments, 5 replicates) were assessed by factorial generalized linear models (GLM) because of the lack of normal distribution for most of the variables according to the Shapiro-Wilk test. The link functions used for each model were selected based on the lowest Akaike information criterion (AIC) selection and characteristics of the data (skewness, counts, continuous variables, proportions) between Gaussian, Gamma, inverse Gaussian, and Tweedie distributions. Differences in treatments were tested using Tukey post-hoc-test with $p <0.05$ as significance criteria. The analyses were conducted in R 4.2.0 \citep{RCoreTeam2018}, and the GLM distributions were extended from the base R core with the Tweedie 2.3.3 package \citep{Dunn2017}. All visualizations were made with the package ggplot2 3.3.3 \citep{Wickham2016}.

@@

Soil pH was significantly affected by the climatic gradient (Figure \cref{fig:soil_properties_panel}), with mean values of 7.7 in PA, 6.2 in SG, 5.9 in LC, and 4.4 in NA, with acidification levels of 6.2 in BSC- to 5.9 in BSC+. In terms of electrical conductivity (EC), a remarkably higher value of \SI{2646.1}{\micro\siemens\per\centi\meter} in PA is outstanding in comparison with the low and homogeneous values of \SI{109.3}{\micro\siemens\per\centi\meter} for SG, \SI{153.8}{\micro\siemens\per\centi\meter} for LC, and \SI{102.3}{\micro\siemens\per\centi\meter} for NA. EC did not differ for the biocrust treatment. Nevertheless, when looking at the site and biocrust in combination, the BSC+ results in a reduction of the EC in PA; meanwhile, in SG, LC, and NA there was no noticeable change.

@@

    % --- Row 4 (2 figures, centered) ---
    % Use slightly wider subfigures here, pushed apart by \hfill
    \begin{subfigure}[b]{0.48\textwidth} % Width suitable for 2 across
        \centering
        \includegraphics[width=\linewidth]{img/M1-Figure_2-10.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \includegraphics[width=\linewidth]{img/M1-Figure_2-11.png}
    \end{subfigure}

    % --- Main Caption and Label for the entire figure ---
    \caption{Mean and standard error of soil properties for Pan de Azúcar (PA), Santa Gracia (SG), La Campana (LC) and Nahuelbuta (NA), without (BSC-) and with biocrust cover (BSC+), and the interaction of both factors. (EC = electrical conductivity, BD = bulk density, C\textsubscript{t} = total carbon, SOC = soil organic carbon, SIC = soil inorganic carbon, N\textsubscript{t} = total nitrogen, CN = carbon to nitrogen ratio). Significant factors include Tukey's Post-hoc-Test indicated by letters on the top of the bars. Different letters represent significant differences at p <0.05.}
    \label{fig:soil_properties_panel} % Use the same descriptive label

@@

Bulk density (BD) showed a significant difference between the study sites, with higher values in the two dryer sites, with \SI{1.5}{\gram\per\cubic\centi\meter} in PA, and \SI{1.6}{\gram\per\cubic\centi\meter} in SG, and a decrease in the more humid sites, with \SI{1.2}{\gram\per\cubic\centi\meter} in LC and \SI{0.6}{\gram\per\cubic\centi\meter} in NA. Biocrust showed no changes for BD in SG and NA, but in PA BSC+ resulted in a reduction of the BD of 18.2\%, while in LC BSC+ showed an increase in BD of 21.9\% (Figure \cref{fig:soil_properties_panel}).

Total carbon (C\textsubscript{t}) content was directly proportional to the humidity along the climatic gradient (Figure \cref{fig:soil_properties_panel}), with values of 1.1\% in PA, 0.8\% in SG, 5.0\% in LC, and 12.5\% in NA, but with a significant decrease when the BSC+ is present, from 5.6\% to 4.2\% in average. Soil inorganic carbon (SIC) was significantly higher in PA, with an average of 0.8\% of the mass of the soil, while in SG, LC, and NA, it was not found. When looking at the SIC along the different sites and under biocrust in combination, the BSC+ results in a reduction of 70.1\% in the SIC for PA, while in SG, LC and NA was no change. Soil organic carbon (SOC) showed a similar pattern as Ct, with a reduction in PA to 0.3\% and the same 0.8\% in SG, 5.0\% in LC, and 12.5\%. Total nitrogen (Nt) content was directly proportional to the humidity along the climatic gradient, with values of 0.04\% for PA, 0.07\% for SG, 0.28\% for LC and 0.51\% for NAError! Reference source not found.). BSC+ showed a reduction of 30.3\% in LC and 22.3\% in NA, while the Nt content remained stable in PA and SG under the influence of BSC. The relation between Ct and Nt, expressed as C/N, showed significantly different values of 33.9 in PA, 12.3 in SG, 16.7 in LC, and 24.5 in NA on average; and with a significant decrease from 27.0 in BSC- to 16.7 in BSC+ (Figure \cref{fig:soil_properties_panel}). It is important to note that although PA and NA present the highest values, the condition changes diametrically when observed together with the BSC+ treatment, with a large dispersion in PA and stable values in NA.

The distribution of the soil particle size classes did not show clear patterns along the climatic gradient, with PA deviating from it in all cases. Despite this, the observed values were significant, with clay values of 9.6\% for PA, 7.3\% in SG, 10.4\% in LC, and 24.6\% in NA, while when looking at the interaction between site and biocrust, there is an increase of 143.4\% in clay content for the BSC+ condition, while for SG, LC and NA was no difference; silt with 28.9\% in PA, 18.7\% in SG, 20.0\% in LC and 21.9\% in NA; and sand with 61.5\% in PA, 73.9\% in SG, 69.6\% in LC and 53.5\% in NA. When biocrusts were present, a significant decrease from 13.6\% to 12.6\% in clays and an increase from 21.2\% to 23.6\% in silt was observed, with higher dispersion for the arid site (PA) (Figure \cref{fig:soil_properties_panel}).

@@

Dry sieving showed a significant difference between study sites but not for the biocrust treatment (Figure 3). Dry aggregates in the 19.0 – 30.0 mm range showed significantly different values of 4.4% in PA, 1.2% in SG, 5.5% in LC, and 36.2% in NA. The fraction 9.5 – 19.0 mm revealed differences in the interaction between study sites and biocrusts, increasing the proportion of aggregates from 13.2% in PA to 19.3% SG and 29.3% in LC, not in NA with only 2.9% in the presence of biocrusts. The aggregates between 6.7 – 9.5 mm showed a significant decrease from 13.3% in PA, 5.2% in SG, 6.5% in LC, and 4.1% in NA. In interaction with biocrusts, it showed a significant increase in the proportion when it was present. The fraction from 4.7 – 6.7mm showed significantly different values of 13.6% in PA, 3.4 % in SG, 5.0% in LC, and 5.6% in NA, while the interaction with biocrusts showed a significant increase when is present (BSC+). Aggregates from 3.4 – 4.7 mm size showed significance among study sites, with 9.8% in PA, 3.6% in SG, 4.4% in LC, and 7.9% in NA. Aggregates between 2.0 – 3.4 mm showed a slightly similar amount of 9.9% in PA, 5.9% in LC, and 14.3% in NA, but with a minor proportion of 4.9% in LC, while when biocrust is present in LC, it showed a slight decrease. Finally, for the dry sieved aggregates under 2 mm, there was a significant reduction for the study sites, with values of 69.1% in SG and 47.3% in LC relative to 30.1% in PA, but not in NA with 28.9%; while the biocrust effect in interaction with the site is significant with 60.6% in SG and 40.8% in LC, indicating the same proportion of aggregates in this site for SG with 30.0% and NA with 39.8%.

In a second stage, aggregate stability under wet conditions was characterized, with a clear difference between sites, while biocrusts had a significant effect only on the aggregate size classes <2.0 mm and 9.5 – 30.0 mm (Figure 3). At the same time, the fraction 19.0 – 30.0 mm showed an increasing significant pattern in the amount of aggregates, with 2.8% in NA, 0.8% in SG, 8.9% in LC, and 29.9% in NA, while biocrusts significantly increased the proportion of aggregates by 35.6%, from 9.0% for BSC- to 12.2% for BSC+. For the fraction 9.5 – 19.0 mm, there were significant differences between the study sites, the biocrust effect, and its interactions, with an increase from 5.4% to 11.7% on average when biocrusts were present. The wet sieved aggregates in the range of 6.7 – 9.5 mm showed significant differences only between the study sites, with 10.6% in PA, 2.5% in SG, 4.2% in LC and 4.3% in NA. Wet aggregates in the range of 4.7 – 6.7 mm showed significant differences between the study sites with 10.6% in PA, 2.3% in SG, 3.7% in LC and 5.8 in NA. The fraction between 3.4 – 4.7 mm showed only significant differences between the study sites, with 6.9% in PA, 2.2% in SG, 3.0% in LC, and 6.1% in NA. Aggregates ranging from 2.0 – 3.4 mm showed differences between the study sites, with 5.8% in PA, 3.2% in SG, 4.2% in LC, and 6.8% in NA. The fractions under 2.0 mm were significantly different for the study sites, with 58.3% in PA, 79.6% in SG, 62.4% in LC, and 41.3% in NA; and for biocrusts from 63.9% for BSC+ treatment to 56.9% for BSC- treatment.

% Placeholder figure 3

When comparing the changes of the aggregate distributions between wet and dry conditions (Figure 4. Mean and standard deviation ), an irregular pattern was observed, with a general decrease in most of the analyzed fractions, except for an increase in the amount of aggregates of 19.0 mm for NA. This was even higher than for the BSC+ treatment. It is important to mention that NA also showed a slight increase in the proportion of 9.5 and 6.7 mm aggregates.

% Placeholder figure 4

Soil aggregate stability was evaluated through different indexes to integrate the different sizes and sieving conditions in a summary value. The difference in mean weight diameter of the aggregates ($\Delta$MWD) showed no significance in any of the conditions (Figure 5). However, there was a significant difference in the difference of the geometric mean diameter of the aggregates ($\Delta$GMD) for the study sites, with values of 1.86 mm for PA, 1.2 mm for SG, 1.4 mm for LC, in comparison with the more stable condition of 0.83 mm for NA. The water-stable aggregates ratio was significant for the study sites, showing differences between NA with 81.1% and the other study sites (57.7% PA, 66.2% SG, and 73.4% LC). The ratio of soil macroaggregates of a diameter less than 2.0 mm (R<2 mm) presents differences in the study sites and for biocrust treatments. SG showed a value of R<2 mm of 79.6% and NA of 41.9%, which was different from PA and LC, with 58.6% and 62.4%, respectively. For biocrust treatments, R<2 mm changed from 63.7% to 57.5% with biocrusts, indicating a biocrust-induced decrease in the proportion for this fraction. Finally, according to the analyzed indexes, NA showed the most stable conditions, and alternating PA and SG showed the most unstable conditions.

% Placeholder figure 5

As shown in Figure 2, the climatic gradient (site) had significant pH, electrical conductivity (EC), bulk density (BD), total carbon (Ct), soil organic carbon (SOC), soil inorganic carbon (SIC), total nitrogen (Nt), C/N, clay, silt, sand, dry and wet aggregates under 30 mm, $\Delta$GMD, WSAR, R<2 mm. Biocrust treatments were significant for clay, silt, pH, EC, Ct, SOC, SIC, Nt, C/N, and wet aggregates from 9.5 to 30.0 mm and >2 mm, and R<2 mm. Finally, the interaction of the site and biocrusts was significant for clay, bulk density (BD), total nitrogen (Nt), dry aggregates from 4.7 to 19.0 mm and 0 to 3.4 mm, wet aggregates from 9.5 to 30.0 mm, and $\Delta$MWD.

@@

The climatic gradient has a significant effect on the stability of soil aggregates. Using the geometric mean diameter ($\Delta$GMD), an index that replaces the linear fitting of $\Delta$MWD with a logarithmic one, significant differences for the study sites along the climatic gradient can be observed (p-value: <0.001). When soil aggregate stability was evaluated with the difference in mean weight diameter ($\Delta$MWD), it did not show significant changes along the climatic gradient. WSAR, an index that shows the ratio of aggregates that persist stably after disruption by water, showed a similar behavior as $\Delta$MWD. The main difference between $\Delta$MWD and $\Delta$GMD is that $\Delta$GMD performs better in non-uniform particulate substances (Hatch and Choate, 1929; Gardner, 1956), which corresponds to soils equilibrated in the content of sand, silt, and clay (Mazurak, 1950) and pointing soil texture indirectly as a critical factor in aggregate stability along the climatic gradient. Further, considering the $\Delta$GMD data, an increase in stability was observed as moving along the climatic gradient to higher water availability conditions except for SG. The lower value of $\Delta$GMD for SG indicates comparably higher aggregate stability as it would be expected when we assume a steady trend from arid to humid climate. In PA, in the drier north, the condition proved to be less stable than NA, despite the high content of inorganic carbon. SG presented the highest ratio of unstable aggregates under the studied range of sizes (highest R<2mm) and NA the lowest, with close to half of it, indicating augmented aggregate stability in the complementary range of sizes.

The effect of the climatic gradient is not only expressed in the stability of soil aggregates, but it is also present with different intensities in a variety of soil properties. The pH decreases continuously from the northern arid to the southern humid study site in accordance with Bernhard et al. (2018). The high pH in PA can be attributed to the constant input of atmospheric aerosols, e.g., salts, gypsum, and calcium carbonates (Ewing et al., 2006) in combination with the arid climate that allows salts to accumulate in the topsoil (Slessarev et al., 2016). Whereas in the southern sites, the forwarding increase of precipitations results in a reduction in the pH due to leaching of soluble salts (Slessarev et al., 2016) and an increase in soil respiration (Orchard and Cook, 1983). The accumulation of soluble salts is well established for the arid site PA, as saline conditions (Allison and Richards, 1954) are indicated by the high electrical conductivity value. These higher amounts of salts have a strong effect in structure degradation dynamics, linked to the destabilizing effect of sodium and stabilizing of carbonates (Corwin, 2021). Although Ct and Nt follow the climatic gradient, when comparing the C/N, PA and NA have higher values. High values of the C/N indicate a nitrogen limitation of plants and other organisms (Brust, 2019), pointing out that this occurs at the two opposite sites along the climatic gradient. This could be explained by the biological activity (Zhang et al., 2013), which may be close to a physiological limitation in the case of NA, while for PA, it may indeed be due to low nitrogen availability (Hooper and Johnson, 1999). However, there was also a high amount of carbonates in PA, which makes PA hardly comparable in terms of Ct. Despite the properties following the climatic gradient, SG deviates from the other sites in terms of higher bulk density (BD), lower clay, and higher sand content. This can indicate a degraded condition for the semi-arid site caused by the current land use, including grazing (Armesto et al., 2007), compacting the surface and thus activating erosive processes (Scholten and Seitz, 2019), in favor of the accumulation of sand particles (Govers, 1985). The aggregate distribution stresses this finding, where SG has a lower proportion of water-stable aggregates >2 mm and a higher water-destabilization of aggregates between 9.5 to 4.7 mm, indicating that the nature of the structuring agent in that zone is water-soluble.

@@

Despite these factors beyond the climatic gradient, biocrusts showed effects on clay, silt, pH, Ct, Nt, C/N, and wet aggregates from 9.5 to 30.0 mm and >2 mm. However, as this was an observational study, it only allows establishing associations between factors and not cause-effect relationships (Cox, 1992; Rosenbaum, 2005). It is thus possible that changes in soil characteristics promote the biocrust establishment, as well as that biocrust establishment triggers changes in these properties (Belnap and Lange, 2003).

>>>>

    $$\Delta MWD = \frac{\sum_{i=1}^{n} X_i*W_i}{\sum_{i=1}^{n} W_i}$$
    where $W_i$ is the corrected mass proportion of stable aggregate fraction $i$ in the total \SIrange[range-phrase=--,range-units=single]{2}{30}{\milli\meter} aggregate fractions, and $X_i$ is the mean diameter of stable aggregate fraction $i$. and $n$ is the number of particle fractions.

@@

The influence of the climatic gradient (study site) and biocrust presence on physicochemical soil parameters and aggregate stability in 40 plots (4 study sites, 2 biocrust treatments, 5 replicates) were assessed by factorial generalized linear models (GLM) because of the lack of normal distribution for most of the variables according to the Shapiro-Wilk test. The link functions used for each model were selected based on the lowest Akaike information criterion (AIC) selection and characteristics of the data (skewness, counts, continuous variables, proportions) between Gaussian, Gamma, inverse Gaussian, and Tweedie distributions. Differences in treatments were tested using Tukey post-hoc-test with $p <0.05$ as significance criteria. The analyses were conducted in R 4.2.0 \citep{RCoreTeam2018}, and the GLM distributions were extended from the base R core with the Tweedie 2.3.3 package \citep{Dunn2017}. All visualizations were made with the package ggplot2 3.3.3 \citep{Wickam2016}.

@@

Soil pH was significantly affected by the climatic gradient (Figure \ref{fig:M1-F2-soil_properties_panel}), with mean values of 7.7 in PA, 6.2 in SG, 5.9 in LC, and 4.4 in NA, with acidification levels of 6.2 in BSC- to 5.9 in BSC+. In terms of electrical conductivity (EC), a remarkably higher value of \SI{2646.1}{\micro\siemens\per\centi\meter} in PA is outstanding in comparison with the low and homogeneous values of \SI{109.3}{\micro\siemens\per\centi\meter} for SG, \SI{153.8}{\micro\siemens\per\centi\meter} for LC, and \SI{102.3}{\micro\siemens\per\centi\meter} for NA. EC did not differ for the biocrust treatment. Nevertheless, when looking at the site and biocrust in combination, the BSC+ results in a reduction of the EC in PA; meanwhile, in SG, LC, and NA there was no noticeable change.

@@

    % --- Row 4 (2 figures, centered) ---
    % Use slightly wider subfigures here, pushed apart by \hfill
    \begin{subfigure}[b]{0.32\textwidth} % Width suitable for 2 across
        \centering
        \includegraphics[width=\linewidth]{img/M1-Figure_2-10.png}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
        \centering
        \includegraphics[width=\linewidth]{img/M1-Figure_2-11.png}
    \end{subfigure}

    % --- Main Caption and Label for the entire figure ---
    \caption{Mean and standard error of soil properties for Pan de Azúcar (PA), Santa Gracia (SG), La Campana (LC) and Nahuelbuta (NA), without (BSC-) and with biocrust cover (BSC+), and the interaction of both factors. (EC = electrical conductivity, BD = bulk density, C\textsubscript{t} = total carbon, SOC = soil organic carbon, SIC = soil inorganic carbon, N\textsubscript{t} = total nitrogen, C/N = carbon to nitrogen ratio). Significant factors include Tukey's Post-hoc-Test indicated by letters on the top of the bars. Different letters represent significant differences at $p <0.05$.}
    \label{fig:M1-F2-soil_properties_panel} % Use the same descriptive label

@@

Bulk density (BD) showed a significant difference between the study sites, with higher values in the two dryer sites, with \SI{1.5}{\gram\per\cubic\centi\meter} in PA, and \SI{1.6}{\gram\per\cubic\centi\meter} in SG, and a decrease in the more humid sites, with \SI{1.2}{\gram\per\cubic\centi\meter} in LC and \SI{0.6}{\gram\per\cubic\centi\meter} in NA. Biocrust showed no changes for BD in SG and NA, but in PA BSC+ resulted in a reduction of the BD of 18.2\%, while in LC BSC+ showed an increase in BD of 21.9\% (Figure \ref{fig:soil_properties_panel}).

Total carbon (C\textsubscript{t}) content was directly proportional to the humidity along the climatic gradient (Figure \ref{fig:soil_properties_panel}), with values of 1.1\% in PA, 0.8\% in SG, 5.0\% in LC, and 12.5\% in NA, but with a significant decrease when the BSC+ is present, from 5.6\% to 4.2\% in average. Soil inorganic carbon (SIC) was significantly higher in PA, with an average of 0.8\% of the mass of the soil, while in SG, LC, and NA, it was not found. When looking at the SIC along the different sites and under biocrust in combination, the BSC+ results in a reduction of 70.1\% in the SIC for PA, while in SG, LC and NA was no change. Soil organic carbon (SOC) showed a similar pattern as C\textsubscript{t}, with a reduction in PA to 0.3\% and the same 0.8\% in SG, 5.0\% in LC, and 12.5\%. Total nitrogen (N\textsubscript{t}) content was directly proportional to the humidity along the climatic gradient, with values of 0.04\% for PA, 0.07\% for SG, 0.28\% for LC and 0.51\% for (Figure \ref{fig:soil_properties_panel}). BSC+ showed a reduction of 30.3\% in LC and 22.3\% in NA, while the N\textsubscript{t} content remained stable in PA and SG under the influence of BSC. The relation between C\textsubscript{t} and N\textsubscript{t}, expressed as C/N, showed significantly different values of 33.9 in PA, 12.3 in SG, 16.7 in LC, and 24.5 in NA on average; and with a significant decrease from 27.0 in BSC- to 16.7 in BSC+ (Figure \ref{fig:soil_properties_panel}). It is important to note that although PA and NA present the highest values, the condition changes diametrically when observed together with the BSC+ treatment, with a large dispersion in PA and stable values in NA.

The distribution of the soil particle size classes did not show clear patterns along the climatic gradient, with PA deviating from it in all cases. Despite this, the observed values were significant, with clay values of 9.6\% for PA, 7.3\% in SG, 10.4\% in LC, and 24.6\% in NA, while when looking at the interaction between site and biocrust, there is an increase of 143.4\% in clay content for the BSC+ condition, while for SG, LC and NA was no difference; silt with 28.9\% in PA, 18.7\% in SG, 20.0\% in LC and 21.9\% in NA; and sand with 61.5\% in PA, 73.9\% in SG, 69.6\% in LC and 53.5\% in NA. When biocrusts were present, a significant decrease from 13.6\% to 12.6\% in clays and an increase from 21.2\% to 23.6\% in silt was observed, with higher dispersion for the arid site (PA) (Figure \ref{fig:soil_properties_panel}).

@@

Dry sieving showed a significant difference between study sites but not for the biocrust treatment (Figure \ref{fig:M1-F3}). Dry aggregates in the \SIrange[range-phrase=--,range-units=single]{19.5}{30.0}{\milli\meter} range showed significantly different values of 4.4\% in PA, 1.2\% in SG, 5.5\% in LC, and 36.2\% in NA. The fraction \SIrange[range-phrase=--,range-units=single]{9.5}{19.0}{\milli\meter} revealed differences in the interaction between study sites and biocrusts, increasing the proportion of aggregates from 13.2\% in PA to 19.3\% SG and 29.3\% in LC, not in NA with only 2.9\% in the presence of biocrusts. The aggregates between \SIrange[range-phrase=--,range-units=single]{6.7}{9.5}{\milli\meter} showed a significant decrease from 13.3\% in PA, 5.2\% in SG, 6.5\% in LC, and 4.1\% in NA. In interaction with biocrusts, it showed a significant increase in the proportion when it was present. The fraction from \SIrange[range-phrase=--,range-units=single]{4.7}{6.7}{\milli\meter} showed significantly different values of 13.6\% in PA, 3.4\% in SG, 5.0\% in LC, and 5.6\% in NA, while the interaction with biocrusts showed a significant increase when is present (BSC+). Aggregates from \SIrange[range-phrase=--,range-units=single]{3.4}{4.7}{\milli\meter} size showed significance among study sites, with 9.8\% in PA, 3.6\% in SG, 4.4\% in LC, and 7.9\% in NA. Aggregates between \SIrange[range-phrase=--,range-units=single]{2.0}{3.4}{\milli\meter} showed a slightly similar amount of 9.9\% in PA, 5.9\% in LC, and 14.3\% in NA, but with a minor proportion of 4.9\% in LC, while when biocrust is present in LC, it showed a slight decrease. Finally, for the dry sieved aggregates under \SI{2}{\milli\meter}, there was a significant reduction for the study sites, with values of 69.1\% in SG and 47.3\% in LC relative to 30.1\% in PA, but not in NA with 28.9\%; while the biocrust effect in interaction with the site is significant with 60.6\% in SG and 40.8\% in LC, indicating the same proportion of aggregates in this site for SG with 30.0\% and NA with 39.8\%.

In a second stage, aggregate stability under wet conditions was characterized, with a clear difference between sites, while biocrusts had a significant effect only on the aggregate size classes $<$\SI{2.0}{\milli\meter} and \SIrange[range-phrase=--,range-units=single]{19.0}{30.0}{\milli\meter} (Figure \ref{fig:M1-F3}). At the same time, the fraction \SIrange[range-phrase=--,range-units=single]{19.0}{30.0}{\milli\meter} showed an increasing significant pattern in the amount of aggregates, with 2.8\% in NA, 0.8\% in SG, 8.9\% in LC, and 29.9\% in NA, while biocrusts significantly increased the proportion of aggregates by 35.6\%, from 9.0\% for BSC- to 12.2\% for BSC+. For the fraction \SIrange[range-phrase=--,range-units=single]{9.5}{19.0}{\milli\meter}, there were significant differences between the study sites, the biocrust effect, and its interactions, with an increase from 5.4\% to 11.7\% on average when biocrusts were present. The wet sieved aggregates in the range of \SIrange[range-phrase=--,range-units=single]{6.7}{9.5}{\milli\meter} showed significant differences only between the study sites, with 10.6\% in PA, 2.5\% in SG, 4.2\% in LC and 4.3\% in NA. Wet aggregates in the range of \SIrange[range-phrase=--,range-units=single]{4.7}{6.7}{\milli\meter} showed significant differences between the study sites with 10.6\% in PA, 2.3\% in SG, 3.7\% in LC and 5.8 in NA. The fraction between \SIrange[range-phrase=--,range-units=single]{3.4}{4.7}{\milli\meter} showed only significant differences between the study sites, with 6.9\% in PA, 2.2\% in SG, 3.0\% in LC, and 6.1\% in NA. Aggregates ranging from \SIrange[range-phrase=--,range-units=single]{2.0}{3.4}{\milli\meter} showed differences between the study sites, with 5.8\% in PA, 3.2\% in SG, 4.2\% in LC, and 6.8\% in NA. The fractions under \SI{2.0}{\milli\meter} were significantly different for the study sites, with 58.3\% in PA, 79.6\% in SG, 62.4\% in LC, and 41.3\% in NA; and for biocrusts from 63.9\% for BSC+ treatment to 56.9\% for BSC- treatment.

\begin{figure}[H]
	\centering
	\includegraphics[width=1\textwidth]{img/M1-Figure_3.png}
	\caption{Mean value and standard error of aggregate proportion for dry (a) and wet (b) sieved fractions for Pan de Azúcar (PA), Santa Gracia (SG), La Campana (LC), and Nahuelbuta (NA) for biocrust (BSC+) and non-biocrust (BSC-) treatments. Significant factors and corresponding models for response variables are included in Appendix \ref{tab:appendix_a}.}
	\label{fig:M1-F3}
\end{figure}

When comparing the changes of the aggregate distributions between wet and dry conditions (Figure \ref{fig:M1-F4}), an irregular pattern was observed, with a general decrease in most of the analyzed fractions, except for an increase in the amount of aggregates of \SI{19.0}{\milli\meter} for NA. This was even higher than for the BSC+ treatment. It is important to mention that NA also showed a slight increase in the proportion of \SI{9.5}{\milli\meter} and \SI{6.7}{\milli\meter} aggregates.

\begin{figure}[H]
	\centering
	\includegraphics[width=1\textwidth]{img/M1-Figure_4.png}
	\caption{Mean and standard deviation of aggregate size distribution changes between dry and wet sieving for the sites Pan de Azúcar (PA), Santa Gracia (SG), La Campana (LC), and Nahuelbuta (NA) for biocrust (BSC+) and non-biocrust (BSC-) treatments. Error bars were, in some cases, truncated for better visualization.}
	\label{fig:M1-F4}
\end{figure}

Soil aggregate stability was evaluated through different indexes to integrate the different sizes and sieving conditions in a summary value. The difference in mean weight diameter of the aggregates ($\Delta$MWD) showed no significance in any of the conditions (Figure \ref{fig:M1-F5}). However, there was a significant difference in the difference of the geometric mean diameter of the aggregates ($\Delta$GMD) for the study sites, with values of \SI{1.86}{\milli\meter} for PA, \SI{1.2}{\milli\meter} for SG, \SI{1.4}{\milli\meter} for LC, in comparison with the more stable condition of \SI{0.83}{\milli\meter} for NA. The water-stable aggregates ratio was significant for the study sites, showing differences between NA with 81.1\% and the other study sites (57.7\% PA, 66.2\% SG, and 73.4\% LC). The ratio of soil macroaggregates of a diameter less than \SI{2.0}{\milli\meter} (R\textsubscript{$<$\SI{1.4}{\milli\meter}}) presents differences in the study sites and for biocrust treatments. SG showed a value of R\textsubscript{<\SI{2}{\milli\meter}} of 79.6\% and NA of 41.9\%, which was different from PA and LC, with 58.6\% and 62.4\%, respectively. For biocrust treatments, R\textsubscript{$<$\SI{1.4}{\milli\meter}} changed from 63.7\% to 57.5\% with biocrusts, indicating a biocrust-induced decrease in the proportion for this fraction. Finally, according to the analyzed indexes, NA showed the most stable conditions, and alternating PA and SG showed the most unstable conditions.

\begin{figure}[H]
	\centering
	\includegraphics[width=1\textwidth]{img/M1-Figure_5.png}
	\caption{Mean and standard error of aggregate stability indexes for Pan de Azúcar (PA), Santa Gracia (SG), La Campana (LC),
    and Nahuelbuta (NA) for biocrust (BSC+) and non-biocrust (BSC-) treatments. $\Delta$MWD=difference in mean weight diameter,
    $\Delta$GMD=difference in geometric mean diameter, WSAR=water stability aggregate ratio, R\textsubscript{<\SI{2}{\milli\meter}}=ratio of aggregates less than \SI{2}{\milli\meter}. Significant factors and corresponding models for response variables are included in Appendix \ref{tab:appendix_a}.}
	\label{fig:M1-F5}
\end{figure}

As shown in Figure \ref{fig:M1-F2-soil_properties_panel}, the climatic gradient (site) had significant pH, electrical conductivity (EC), bulk density (BD), total carbon (C\textsubscript{t}), soil organic carbon (SOC), soil inorganic carbon (SIC), total nitrogen (N\textsubscript{t}), C/N, clay, silt, sand, dry and wet aggregates under \SI{30}{\milli\meter}, $\Delta$GMD, WSAR, R\textsubscript{<\SI{2}{\milli\meter}}. Biocrust treatments were significant for clay, silt, pH, EC, C\textsubscript{t}, SOC, SIC, N\textsubscript{t}, C/N, and wet aggregates from \SIrange{9.5}{30.0}{\milli\meter} and $>$\SI{2}{\milli\meter}, and R\textsubscript{<\SI{2}{\milli\meter}}. Finally, the interaction of the site and biocrusts was significant for clay, bulk density (BD), total nitrogen (N\textsubscript{t}), dry aggregates from \SIrange{4.7}{19.0}{\milli\meter} and \SIrange{0}{3.4}{\milli\meter}, wet aggregates from \SIrange{9.5}{30.0}{\milli\meter}, and $\Delta$MWD.

@@

The climatic gradient has a significant effect on the stability of soil aggregates. Using the geometric mean diameter ($\Delta$GMD), an index that replaces the linear fitting of $\Delta$MWD with a logarithmic one, significant differences for the study sites along the climatic gradient can be observed ($p-value: <0.001$). When soil aggregate stability was evaluated with the difference in mean weight diameter ($\Delta$MWD), it did not show significant changes along the climatic gradient. WSAR, an index that shows the ratio of aggregates that persist stably after disruption by water, showed a similar behavior as $\Delta$MWD. The main difference between $\Delta$MWD and $\Delta$GMD is that $\Delta$GMD performs better in non-uniform particulate substances \citep{Hatch1929,Gardner1956}, which corresponds to soils equilibrated in the content of sand, silt, and clay \citep{Mazurak1950} and pointing soil texture indirectly as a critical factor in aggregate stability along the climatic gradient. Further, considering the $\Delta$GMD data, an increase in stability was observed as moving along the climatic gradient to higher water availability conditions except for SG. The lower value of $\Delta$GMD for SG indicates comparably higher aggregate stability as it would be expected when we assume a steady trend from arid to humid climate. In PA, in the drier north, the condition proved to be less stable than NA, despite the high content of inorganic carbon. SG presented the highest ratio of unstable aggregates under the studied range of sizes (highest R\textsubscript{<\SI{2}{\milli\meter}}) and NA the lowest, with close to half of it, indicating augmented aggregate stability in the complementary range of sizes.

The effect of the climatic gradient is not only expressed in the stability of soil aggregates, but it is also present with different intensities in a variety of soil properties. The pH decreases continuously from the northern arid to the southern humid study site in accordance with \citet{Bernhard2018}. The high pH in PA can be attributed to the constant input of atmospheric aerosols, e.g., salts, gypsum, and calcium carbonates \citep{Ewing2006} in combination with the arid climate that allows salts to accumulate in the topsoil \citep{Slessarev2016}. Whereas in the southern sites, the forwarding increase of precipitations results in a reduction in the pH due to leaching of soluble salts \citep{Slessarev2016} and an increase in soil respiration \citep{Orchard1983}. The accumulation of soluble salts is well established for the arid site PA, as saline conditions \citep{Allison1954} are indicated by the high electrical conductivity value. These higher amounts of salts have a strong effect in structure degradation dynamics, linked to the destabilizing effect of sodium and stabilizing of carbonates \citep{Corwin2021}. Although C\textsubscript{t} and N\textsubscript{t} follow the climatic gradient, when comparing the C/N, PA and NA have higher values. High values of the C/N indicate a nitrogen limitation of plants and other organisms \citep{Brust2019}, pointing out that this occurs at the two opposite sites along the climatic gradient. This could be explained by the biological activity \citep{Zhang2013}, which may be close to a physiological limitation in the case of NA, while for PA, it may indeed be due to low nitrogen availability \citep{Hooper1999}. However, there was also a high amount of carbonates in PA, which makes PA hardly comparable in terms of C\textsubscript{t}. Despite the properties following the climatic gradient, SG deviates from the other sites in terms of higher bulk density (BD), lower clay, and higher sand content. This can indicate a degraded condition for the semi-arid site caused by the current land use, including grazing \citep{Armesto2007}, compacting the surface and thus activating erosive processes \citep{Scholten2019}, in favor of the accumulation of sand particles \citep{Govers1985}. The aggregate distribution stresses this finding, where SG has a lower proportion of water-stable aggregates $>$\SI{2}{\milli\meter} and a higher water-destabilization of aggregates between \SIrange{9.5}{4.7}{\milli\meter}, indicating that the nature of the structuring agent in that zone is water-soluble.

@@

Despite these factors beyond the climatic gradient, biocrusts showed effects on clay, silt, pH, C\textsubscript{t}, N\textsubscript{t}, C/N, and wet aggregates from \SIrange{9.5}{30.0}{\milli\meter} and $>$\SI{2}{\milli\meter}. However, as this was an observational study, it only allows establishing associations between factors and not cause-effect relationships \citep{Cox1992,Rosenbaum2005}. It is thus possible that changes in soil characteristics promote the biocrust establishment, as well as that biocrust establishment triggers changes in these properties \citep{Belnap2003}.
